Using Random Walk Simulations to Calculate Ground State Energies in Quantum Physics

### 2020-04-19 Paper/main.tex
Rewrote Theoretical Background section

% author names 
\author{Sai Pandian, ID: 29899923}% <-this % stops a space

@@

\begin{IEEEkeywords}
keywords, temperature, xxxx equation, etc.
\end{IEEEkeywords}

\section{Introduction}
% Here we have the typical use of a "W" for an initial drop letter
% and "RITE" in caps to complete the first word.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an issue)

\IEEEPARstart{R}{andom} walks are a simple model of motion in which a path is
described as a succession of steps with each step being in a random direction on
some vector space. Random walks are useful in modelling many physical processes
that are perceived as having a pseudo-random mechanism \todo{cite}. Their
applications range from Biology to the Stock Market. This paper aims to explore
how Random Walk Simulations can be implemented to calculate the ground state
energies of different quantum systems. \todo{more}

@@

Random walks work descibe motion as a series of steps, with the direction of
each step being random. Figure \ref{fig:cartoon} demonstrates this concept by
considering a one-dimensional line. On this line, the ``walker'' starts at the
origin, and walks a total of 20 steps. The walker at each step has a equal
chance of stepping forwards, or stepping backwards.

\begin{figure}%[H]%%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{Illustrations, graphs, and photographs may fit across both columns,
      if necessary. Your artwork must be in place in the article.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

As can be seen from the figure, it is not uncommon for the average position of
the walker to be away from the origin once the walk has concluded. This is
because the position of the walker at any step is dependent only on the position
at the previous step. If the walker happens to take a number of steps in the
same direction, it is highly unlikely for the walker to take an equal number of
steps in the opposite direction, in order to return to the origin. Thus, in such
a case, it is likely that the walker will conclude the walk in a position away
from the origin.

In this example, the walker had an equal likelihood of stepping forward or
backwards, but we can bias in one way if we want to model a particular
phenomenon. The walk is also easily extensible to multiple dimensions, as for
each extra dimension, one extra random number must be drawn. This gives a linear
increase in time-complexity, which is convenient for long computations.

Random Walk simulations can be applied to calculate the ground state energies of
Quantum Systems by solving the Schr\"{o}dinger equation \todo{cite}. Consider a
one-dimensional simple lattice system. A walker can start at a position labelled
the origin, and then proceed in a random walk. The random walk concludes after a
certain number of steps or if the walker ``dies'' on a particular step. The
probability that the walker will die is dependent on a defined potential at that
point. For example, in an infinite square well system, the potential is 0 within
a well, and infinite outside the well. So the walker will have no chance of
dying whilst inside the well, but will die if they step outside.

Let us define $p(j,n)$ as the probability that the walker will be at position
$j$ after $n$ steps and $a(j)$ as the probability that the walker will die at
position $j$. If we say:

\begin{equation}
  p(j,n) = q(j) \exp(-\lambda n)
  \label{eq:firsteq}
\end{equation}
where $q(j)$ is an arbitrary function and $\lambda$ is the rate of death, then:
\begin{equation}
  -\frac{1}{2} \frac{d^2q(j)}{dj^2} + a(j)q(j) = \lambda q(j)
  \label{eq:randomwave}
\end{equation}

The full derivation is taken from \todo{cite} and is shown in Appendix
\ref{appendix:derivation}.

Notice that Equation \ref{eq:randomwave} is in the form of a wave equation,
specifically the Schr\"{o}dinger equation:
\begin{equation}
  \label{eq:schrodinger}
  \frac{-\hbar^2}{2m}\frac{d^2 \psi}{dr^2} + V(r)\psi(r) = E\psi(r)
\end{equation}
Comparing Equation \ref{eq:randomwave} and Equation \ref{eq:schrodinger}, we can
see that the probability of death $a(j)$ is related to the potential $V(r)$,
and the rate of death $\lambda$ is related to the Energy $E$. Thus, by running
the simulation for a particular potential (i.e. rate of death), one can
determine the ground state energy of a system by calculating the rate of
death. This is the technique employed in this paper.

@@

The Random Walk is implemented computationally \todo{cite} and is extensible
to multiple dimensions. For a $N$-dimensional lattice, at each point there are
$2N$ possible directions for the walker to move. So a direction is selected
randomly using a Mersenne Twister pseudorandom number generator (pRNG). This algorithm
was chosen as it is a very fast implementation of a pRNG \todo{cite} and it
passes many statistical randomness tests \todo{cite}, having a very large
period of $2^{19937-1}$.

@@

If we compare this to Equation \ref{eq:randomwave}, allowing $a(j) = 0$, we see
that:
\begin{equation}
  \begin{split}
    \lambda &= \frac{mJ^2E}{\hbar^2R^2} \\
    E &= \lambda R^2 \frac{\hbar^2}{mJ^2}
  \end{split}
  \nonumber
\end{equation}

@@

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
      particular step in an infinite square well with $J = 20$. This is a
      cumulative sum of the bins in Figure \ref{fig:hist}. As can be seen, the
      peak is at the origin, since it is certain that no walker will die on the
      first step. An exponential decay curve provides a good fit, with an $R^2$
      value of 0.9999998, very close to 1. Plotting on a logarithmic scale shows
      the straight line fit, the gradient of which is $\lambda$, giving $\lambda
\todo{fix centering of cum plot}
      J^2 = 1.2334$, with a negligible error.}
  \label{fig:cumplots}
\end{figure*}

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random Walk computational simulations. The
investigation focussed on an Infinite Square Well, but is easily extended to a
Symmetric potential. The simualation was written in C++ to maximise performance
and is available at \todo{cite}.

@@

This method provided $\lambda J^2 = 1.2339 \pm 0.0008$, which is very close to
the theoretical result.

@@

\appendices
\section{Random Walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Let us denote the probability that the walker is at position $j$ after $n$ steps
as $p(j,n)$ and the probabilty that the walker will die at position $j$ as
$a(j)$. We will denote the position-step number coordinates as $[j, n]$. Thus,
for a walker to have coordinates $[j, n+1]$, in the previous step, they must have
had coordinates $[j \pm 1, n]$. So we can say:

>>>>

% author names 
\author{Sai Pandian, ID: 29899923}%

@@

\section{Introduction}
\IEEEPARstart{T}{he} developments in computer power in recent years has opened
up the possibility for novel computational solutions to many problems that were
otherwise limited. We now have access to computers that can numerically solve
problems and run simulations that were previously too computationally intensive
to do.

One such problem is the calculation of the ground state energies of different
quantum systems. Many methods exist for solving these problems, using both
classical and quantum computational approaches. Whilst generally it may be true
that Quantum computating approahes are more suited to solving such problems
\cite{Mazzola}, these approaches are still in infancy and are not readily
accessible. Thus, there exist an abundance of classical computing approaches,
such as the Variational-Relaxiation algorithm \cite{Schroeder2017}. This is a
particularly useful algorithm for two-dimensional systems with non-separable
potentials but quickly becomes difficult to implement for more complex systems.

In this investigation, I explore an alternative Monte Carlo method using Random
Walk simulations, and demonstrate its efficacy with simple systems.

@@

\subsection{Random Walk Simulations}

Random walks descibe motion as a series of steps in some vector space, with the
direction of each step being random, i.e. with there being an equal likelihood
of a step in each possible direction. Consider a simple 1-dimensional line and a
walker standing at a point labelled the origin. The walker takes 20 steps, the
direction of each step (forwards or backwards on the line) is random. The
position of the walker at each step is shown in Figure \ref{fig:cartoon}.

\begin{figure}%[H]%%[!ht]
  \begin{center}
    \includegraphics[width=0.45\textwidth]{images/cartoon.pdf}
    \caption{The position of a walker at each step in a 20-step random walk
      simulation. The average position of the walker does not necessarily remain
      at the origin as one might naively expect. Instead it is normal for the
      walker to finish the walk at some distance from the origin, as shown here.}
    \label{fig:cartoon}
  \end{center}
\end{figure}

As we can see from the figure, it is not uncommon for the average position of
the walker to be away from the origin once the walk has concluded. This is
because the position of the walker at any step is dependent only on the position
at the previous step. If the walker happens to take a number of steps in the
same direction, it is highly unlikely for the walker to take an equal number of
steps in the opposite direction, in order to return to the origin. Thus, in such
a case, it is likely that the walker will conclude the walk in a position away
from the origin.

In this example, the walker had an equal likelihood of stepping forward or
backwards, but we can bias in one way if we want to model a particular
phenomenon. The walk is also easily extensible to multiple dimensions, as for
each extra dimension, one extra random number must be drawn. This gives a linear
increase in time-complexity, which is convenient for more complex systems.

\subsection{Application to Quantum Systems}

This investigation applied Random walk simulations to calculate the ground state
energies of Quantum Systems by solving the Schr\"{o}dinger equation, as shown in
\todo{cite}.

We can extend the simple one-dimensional line model from Figure
\ref{fig:cartoon} to a quantum system by imagining it as a one-dimensional
simple lattice system. A walker can start at a position labelled the origin, and
then proceed in a random walk. The random walk concludes if the walker ``dies''
on a particular step. The probability that the walker will die is dependent on a
defined potential at that point. For example, in an infinite square well system,
the potential is 0 within a well, and infinite outside the well. So the walker
will have no chance of dying whilst inside the well, but will die if they step
outside.

Let us define $p(j,n)$ as the probability that the walker will be at position
$j$ after $n$ steps and $a(j)$ as the probability that the walker will die at
position $j$. Thus, for this to be the case, the walker would have had position
$j\pm1$ after $n-1$ steps. Thus, we can say:

\begin{equation}
  p(j, n+1) =  \frac{1}{2}(1-a(j))(p(j-1,n) + p(j+1,n))
  \nonumber
\end{equation}

If $a(j)$ is small and $n$ is large, we can use Taylor's expansion to expand
each term in this expression and obtain a differential equation. If we then
make the ansatz substitution:

\begin{equation}
  p(j,n) = q(j) \exp(-\lambda n)
  \label{eq:firsteq}
\end{equation}
where $q(j)$ is an arbitrary function and $\lambda$ is the rate of death, then
we find:

\begin{equation}
  -\frac{1}{2} \frac{d^2q(j)}{dj^2} + a(j)q(j) = \lambda q(j)
  \label{eq:randomwave}
\end{equation}
The full derivation is taken from \cite{MarcusNewton2020} and is shown in
Appendix \ref{appendix:derivation}. Notice that Equation \ref{eq:randomwave} is
in the form of the Schr\"{o}dinger equation:

\begin{equation}
  \label{eq:schrodinger}
  \frac{-\hbar^2}{2m}\frac{d^2 \psi}{dr^2} + V(r)\psi(r) = E\psi(r)
\end{equation}
Comparing Equation \ref{eq:randomwave} and Equation \ref{eq:schrodinger}, we can
see that the probability of death $a(j)$ is related to the potential $V(r)$,
and the rate of death $\lambda$ is related to the Energy $E$.

We can extend this to an infinite square well by choosing our probability of
death such that:

\begin{equation}
  \label{eq:squarewell}
    a(j) =
    \begin{cases}
      0,& \text{if } j < J\\
      \infty,& \text{if } j \geq J
    \end{cases}
    \nonumber
\end{equation}
where J is a well-defined boundary. So the walker cannot die inside the well but
wil immediately die when they step outside, which we expect will definitely
happen given sufficient time. Inside the well, the potential is simply that of a
free particle:

\begin{equation}
  \frac{-\hbar^2}{2m}\frac{d^2 \psi}{dx^2} = E\psi(x)
  \nonumber
\end{equation}
where $x$ is the position of the walker. By making the substitution
$x=J\frac{j}{R}$, such that $j=R$ corresponds to the walker being at the
boundary, we find:

\begin{equation}
  \begin{split}
    -\frac{\hbar^2}{2m}\frac{R^2}{J^2}\frac{d^2\psi}{dj^2} &= E\psi(x)\\
    -\frac{1}{2}\frac{d^2\psi}{dj^2} &= \frac{mJ^2E}{\hbar^2R^2}\psi(x)
  \end{split}
  \nonumber
\end{equation}
If we compare this to Equation \ref{eq:randomwave}, allowing $a(j) = 0$ inside
the well, we see that:

\begin{equation}
  E = \lambda R^2 \frac{\hbar^2}{mJ^2}
  \nonumber
\end{equation}
where only $\lambda$ is an unknown quantity. Thus, by calculating rate of death
$\lambda$, we can calculate the ground state energy of the infinite square
well. The known ground state energy of an infnite square well is
$\frac{\pi^2}{8}\frac{\hbar^2}{mJ^2}$ \todo{cite}, so we expect to find $\lambda
R^2$ close to $\frac{\pi}{8}$.

@@

The Random walk is implemented computationally \todo{cite} and is extensible
to multiple dimensions. For a $N$-dimensional lattice, at each point there are
$2N$ possible directions for the walker to move. So a direction is selected
randomly using a Mersenne Twister pseudorandom number generator (pRNG). This algorithm
was chosen as it is a very fast implementation of a pRNG \todo{cite} and it
passes many statistical randomness tests \todo{cite}, and has a very large
period of $2^{19937-1}$.

@@

If we compare this to Equation \ref{eq:randomwave}, allowing $a(j) = 0$ inside
the well, we see that:
\begin{equation}
  \begin{split}
    \lambda &= \frac{mJ^2E}{\hbar^2R^2} \\
    E &= \lambda R^2 \frac{\hbar^2}{mJ^2}
  \end{split}
  \nonumber
\end{equation}

@@

\begin{figure*}%[ht!]
  \centering
  \subfloat[Histogram of probability of surviving to at least a particlar step]
  {\includegraphics[width=0.5\textwidth]{images/cum_exp_plot.pdf}
  \label{fig:cumhist}}
  \centering
  \subfloat[Histogram of probability of surviving to at least a particular step
  in a logarithmic scale]
  {\includegraphics[width=0.5\textwidth]{images/cum_line_plot.pdf}
  \label{fig:cumloghist}}
  \caption{Histogram showing the Probability of surviving to at least a
      particular step in an infinite square well with $J = 20$. This is a
      cumulative sum of the bins in Figure \ref{fig:hist}. As can be seen, the
      peak is at the origin, since it is certain that no walker will die on the
      first step. An exponential decay curve provides a good fit, with an $R^2$
      value of 0.9999998, very close to 1. Plotting on a logarithmic scale shows
      the straight line fit, the gradient of which is $\lambda$, giving $\lambda
      J^2 = 1.2334$, with a negligible error.}
  \label{fig:cumplots}
\end{figure*}

\todo{fix centering of cum plot}

@@

In this investigation we demonstrated a method of calculating ground state
energies of a quantum system using Random walk computational simulations. The
investigation focussed on an Infinite Square Well, but is easily extended to a
Symmetric potential. The simualation was written in C++ to maximise performance
and is available at \todo{cite}.

@@

This method provided $\lambda J^2 = 1.2339 \pm 0.0008$, which lies within the
range of the theoretical result.

@@

\appendices
\section{Random walk Schr\"{o}dinger Equation Derivation}
\label{appendix:derivation}
Let us denote the probability that the walker is at position $j$ after $n$ steps
as $p(j,n)$ and the probabilty that the walker will die at position $j$ as
$a(j)$. We will denote the position-step number coordinates as $[j, n]$. Thus,
for a walker to have coordinates $[j, n+1]$, in the previous step, they must have
had coordinates $[j \pm 1, n]$. So we can say:
